Disentangling the spinal mechanisms of illusory heat and burning sensations in the Thermal Grill Illusion

### 2023-07-14 Manuscript.Rmd
updates

---
title: "Disentangling the spinal mechanisms of illusory heat and burning sensations in the Thermal Grill Illusion"
author: 
  - name : "Alexandra G. Mitchell"
    affiliation : "1"
  - name : "Jesper Fischer Ehmsen"
    affiliation : "1"
  - name : "Daniel Elmstrom Christensen"
    affiliation : "2"
  - name : "Anna Villaume Stuckert"
    affiliation : "3"
  - name : "Patrick Haggard"
    affiliation : "4"  
  - name : "Francesca Fardo"
    affiliation : "1"

affiliation:
  - id : "1"
    institution : "Center of Functionally Integrative Neuroscience (CFIN), Department of Clinical Medicine, Aarhus University"
  - id : "2"
    institution : "Interacting Minds Center (IMC), Department XXX, Aarhus University"
date: "`r Sys.Date()`"
output:
  word_document:
  html_document: default
  pdf_document:
    latex_engine: xelatex
bibliography: knitting files/Citations.bib
link-citations: yes
linkcolor: blue
always_allow_html: yes
editor_options: 
  markdown: 
    wrap: sentence
---

@@

The Thermal Grill Illusion (TGI) is a phenomenon in which the juxtaposition of innocuous warm and cold temperatures elicits a burning sensation, offering a unique window to understand how pain can be perceived in response to harmless stimuli.
Much debate has revolved around whether spinal mechanisms are involved in the generation of illusory pain, beyond supraspinal mechanisms.
In this study, we investigated the role of the spinal cord in the generation of the TGI, in two independent experiments, involving a total of 80 healthy individuals.
We applied heat and cold stimuli on dermatomes, namely areas of skin innervated by a single spinal nerve, mapped onto adjacent or nonadjacent spinal segments.
Participants were asked to rate their perceptions of cold, warm, and burning sensations in response to TGI and control stimuli.
Our aims were to investigate thermosensory and painful perceptual components of the TGI, as well as spatial features of the illusion that may illuminate processes underlying thermosensory integration in the spinal cord.
Our findings revealed that both thermosensory and painful components of TGI perception were modulated similarly, with enhanced warm and burning ratings observed when cold and warm stimuli were confined within the same dermatome.
Further, we found no perceptual differences based on the proximal-distal location of the cold stimulus within a single dermatome, but a notable heat enhancement when the cold stimulus was associated with a more caudal segmental location compared to the warm stimulus.
These results provide insights into the organisation of the spinal cord in relation to the thermosensory integration and generation of the TGI.

@@

The thermal grill illusion (TGI) is a perceptual phenomenon that challenges conventional understanding of pain perception.
It is a sensation of burning heat or pain when harmless cold and warm temperatures are applied to the skin simultaneously [@craig_thermal_1994; @craig_functional_1996].
Despite cold and warm temperatures being individually innocuous, their combination produces a contradictory burning sensation, even so the temperatures are insufficient to activate peripheral nociceptors.
The generation of this illusion is thus attributed to central nervous system mechanisms [@craig_thermal_1994; @fardo_beyond_2020].
Recent studies have highlighted the involvement of the spinal cord as an initial site contributing to the TGI [@fardo_organization_2018; @harper_conditioned_2017].
However, the precise mechanisms underpinning integration of cold and warm thermal afferents in the spinal cord, alongside those responsible for the distinctive burning quality to this illusion, are yet to be elucidated.

The TGI is often described as encompassing two distinct perceptual components - an illusion of heat and an illusion of pain [@defrin_spatial_2008, @fardo_beyond_2020]. The illusion of heat, also known as synthetic heat [@fruhstorfer_significance_2003; @green_localization_1977; -@green_synthetic_2002; -@green_temperature_2004], refers to non-painful sensations evoked by the thermal grill [@defrin_spatial_2008; @kern_pharmacological_2008; -@kern_effects_2008; @bouhassira_investigation_2005; @adam_relationships_2014].
The illusion of pain, which is the most recognised aspect of the TGI, is the distinctive burning sensation that accompanies the simultaneous presentation of cold and warm stimuli [@craig_thermal_1994; -@craig_functional_1996; @bach_thermal_2011].
The hallmark of both illusory components is a qualitative change in perception when the cold and warm stimuli are applied concurrently, compared to when they are presented individually.
Historically, the thermosensory and painful components of the TGI were explained by distinct spinal and supraspinal mechanisms, respectively.
The enhanced perception of heat in TGI was explained by a spinal inhibitory mechanism, drawing from observations in an animal model where simultaneous application of cold and warm temperatures reduced the activity of cold-specific spinal neurons compared to when cold was applied alone [@craig_thermal_1994]. Instead, the illusory pain component was ascribed to a disinhibition mechanism at the level of the thalamus, primarily based on the observations of unremitting pain following a thalamic lesion [@craig_thermal_1994; -@bud_craig_new_1998].

Recent human studies on TGI provided differing interpretations of the spinal or supraspinal origin of the illusion.
Two studies posited that the illusory pain of the TGI depends uniquely on supraspinal mechanisms.
This interpretation was based on the observed modulation of the illusion in accordance with a spatiotopic rather than somatotopic representation of the body [@marotta_transforming_2015].
Further, the illusion remained unaltered during concomitant tactile stimulation, suggesting ineffectiveness of tactile gating - a spinally-mediated process involving inhibition of nociceptive activity by concurrent somatosensory activity [@ferre_ineffectiveness_2018].
Counter to this perspective, other research endorsed a spinal contribution to the TGI.
These studies demonstrated that the illusion varied depending on whether cold and warm stimuli were applied to dermatomes mapped either onto adjacent or non-adjacent spinal segments [@fardo_organization_2018].
Participants perceived the stimulation more veridically, consistently with a reduction in TGI perception, when warm or cold stimuli triggered more widespread activity along the spinal cord, corroborating the hypothesis that the spinal cord is an initial site of thermosensory integration underlying TGI.
Further support for spinal mechanisms comes from research demonstrating that both noxious heat and the TGI were comparably reduced by conditioned pain modulation in humans.
This suggests a similar influence of descending modulation, irrespective of whether the painful sensation was triggered by potentially harmful (noxious) or harmless (innocuous) stimuli [@harper_conditioned_2017].
These findings collectively challenge a purely supraspinal hypothesis of the painful component of the TGI and indicate the relevance of spinal mechanisms in the manifestation of both illusory heat and pain within the TGI.

In this paper our objective was twofold.
Firstly, we directly investigated the hypothesis that thermosensory and burning components of the TGI experience are mediated by spinal mechanisms in humans, by manipulating the location of the stimuli within and across dermatomes.
Cold and warm stimuli were presented at a fixed distance on the skin, but depending on their longitudinal or tangential orientation on the arm, they elicited neural activity in a differing number of spinal segments.
Our assumption was that cold and warm-related neural activity in the spinal cord was more focal, when the stimuli were presented within the same dermatome, while more widespread, when the stimuli mapped on spinal segments that were further apart.
Our past work using a similar dermatome manipulation involved measuring the experience of the TGI using a temperature matching task [@fardo_organization_2018], which provides a composite measure of TGI perception, reflecting both thermosensory and burning components.
Here, to probe possible distinctions between the two qualitative components of the TGI, we measured subjective indices of TGI perception using three independent visual analog scale (VAS) ratings of perceived cold, warm and burning sensations.
Secondly, we investigated spatial order effects associated with the integration of cold and warm sensory information at the dermatome (skin) and segmental (i.,e., spine) levels.
At the dermatome level, we used body-related coordinates to define proximal (towards the elbow) and distal (towards the wrist) locations.
At the spinal level, we used segment-related coordinates to define more rostral (towards the head) and more caudal (towards the lower back) locations.
Given the organisation of the spinal cord along a rostral-caudal axis, where each dermatome is represented across multiple spinal segments through the Lissauer tract, this study aimed to glean indirect insights into the spinal mechanisms underpinning thermosensory integration and the generation of the TGI.

@@

The Thermal Grill Illusion (TGI) is characterised by two key phenomena: thermosensory enhancement and illusory pain.
Thermosensory enhancement refers to an amplified perception of heat or cold, when cold and warm stimuli are simultaneously applied, as opposed to when each stimulus is presented individually or paired with a neutral temperature.
Notably, the majority of individuals experience an intensification of heat rather than cold [@fardo_beyond_2020].
Illusory pain, on the other hand, denotes the perception of a burning sensation elicited by the pairing of warm and cold stimuli, an experience that is largely absent or significantly diminished when each stimulus is presented alone or combined with a neutral temperature.
Thus, indicators of a stronger TGI are heightened warm and burning ratings, coupled with reduced cold ratings.
To investigate thermosensory and burning components of the TGI, participants received pairs of temperatures on their forearms and were asked to quantify the levels of cold, warmth, and burning they experienced during each stimulation.
These stimuli consisted of either cold-warm pairs (TGI stimuli), which potentially evoked an illusion of heat and pain, or control stimuli that involved pairing a cold or a warm stimulus of identical temperature as in the TGI condition with a baseline temperature of 30°C (non-TGI stimuli).
All stimulation pairs were presented at a fixed distance on the skin, either within the same dermatome or across dermatomes that mapped onto non-adjacent spinal segments.
For each stimulation, participants reported their ratings using three sequential VAS scales ranging from 0, indicating the lack of a sensation, to 100, indicating an extreme sensation.
For each instance of stimulation, participants were guided to focus their reporting on the sensations originating from a specific thermode.
While the participants were unaware, this location corresponded to either the colder stimulus (Exp. 1) or the warmer stimulus (Exp. 2) of the paired temperatures.
VAS ratings were analysed using zero-inflated beta regressions.

@@

```{r, warning = F, message = F, echo = F,include=FALSE}
if (!require("pacman")) install.packages("pacman")
  pacman::p_load(boot, broom, broom.mixed, brms, car, caret, corrplot, DHARMa, gamlss,
                 ggeffects, gghalves, ggpol, ggpubr, glmmTMB, groupdata2, lme4, lmerTest, 
                 modelsummary, MuMIn, polycor, RColorBrewer, rcompanion, reshape2, Rmisc,
                 rsample, rstatix, simr, tibble, tidyverse, wesanderson, patchwork, cowplot
                 )
source(here::here("scripts","utility.R"))

@@

In keeping with the unique heat and burning features of TGI, our results exhibited a more robust TGI when stimuli were confined within dermatomes compared to when applied across dermatomes, corresponding to non-adjacent spinal segments (Fig. 2).
When rating the cold thermode (Exp. 1), participants reported a significantly reduced subjective experience of cold ($\beta$ = `r stats_model_cold_exp1$beta[2]`, `r make_pvalue(stats_model_cold_exp1$p[2])`), alongside a markedly enhanced subjective experience of warmth ($\beta$ = `r stats_model_warm_exp1$beta[2]`, `r make_pvalue(stats_model_warm_exp1$p[2])`) for TGI stimuli applied within a dermatome compared to across dermatomes.
Burning sensations were enhanced for TGI compared to non-TGI stimuli, but the modulation of the burning ratings by the dermatome condition did not cross the significance threshold ($\beta$ = `r stats_model_burn_exp1$beta[2]`, `r make_pvalue(stats_model_burn_exp1$p[2])`).
In contrast, when rating the warm thermode (Exp. 2), participants reported markedly enhanced burning sensations for TGI stimuli applied within a dermatome compared to across dermatomes ($\beta$ = `r stats_model_burn_exp2$beta[2]`, `r make_pvalue(stats_model_burn_exp2$p[2])`).
However, we did not observe a modulation of cold ($\beta$ = `r stats_model_cold_exp2$beta[2]`, `r make_pvalue(stats_model_cold_exp2$p[2])`) and warm ($\beta$ = `r stats_model_warm_exp2$beta[2]`, `r make_pvalue(stats_model_warm_exp2$p[2])`) thermosensory ratings as in the first experiment.

The findings suggest that the thermosensory quality of the reference thermode (cold or warm) showed differential sensitivity to thermosensory and painful aspects of the TGI experience, and collectively suggested that both the thermosensory enhancement and illusory pain components of the illusion are modulated at the spinal cord level.
This interpretation is consistent with a previous study using a similar dermatome manipulation, but a distinct method to quantify TGI perception [@fardo_organization_2018], as well as another study showing modulation of heat and pain ratings of TGI stimuli by conditioned pain modulation [@harper_conditioned_2017].
All together these results support the role of spinal processes in the generation of distinct perceptual aspects of the TGI.

@@

Previous research demonstrated a phenomenon known as distal inhibition, wherein heat pain ratings tend to increase when a participant evaluates a more distal compared to a more proximal stimulus among two warm stimuli presented on the forearm [@quevedo_psychophysicshyperalgesia_2004; @quevedo_illusion_2007].
In our study, we sought to investigate whether this effect of distal inhibition would also influence TGI perception within a single dermatome (Fig. 3).
Firstly the cold intensity was less when the cold probe was located more distally for both control and TGI trials (TGI: $\beta$ = `r stats_within_cold_exp1_TGI$beta[2]`, `r make_pvalue(stats_within_cold_exp1_TGI$p[2])`; control: $\beta$ = `r stats_within_cold_exp1_control$beta[2]`, `r make_pvalue(stats_within_cold_exp1_control$p[2])`) with no significant interaction between them $\beta$ = `r stats_within_cold_exp1_control$beta[5]`, `r make_pvalue(stats_within_cold_exp1_control$p[5])`.
No difference in intensity was found for the warm ratings (TGI: $\beta$ = `r stats_within_warm_exp1_TGI$beta[2]`, `r make_pvalue(stats_within_warm_exp1_TGI$p[2])`; control: $\beta$ = `r stats_within_warm_exp1_control$beta[2]`, `r make_pvalue(stats_within_warm_exp1_control$p[2])`) with no significant interaction between them $\beta$ = `r stats_within_warm_exp1_control$beta[5]`, `r make_pvalue(stats_within_warm_exp1_control$p[5])`.
significant differences was also found for the burning ratings indicating more distally located cold probe decreased ratings.
(TGI: $\beta$ = `r stats_within_burn_exp1_TGI$beta[2]`, `r make_pvalue(stats_within_burn_exp1_TGI$p[2])`; control: $\beta$ = `r stats_within_burn_exp1_control$beta[2]`, `r make_pvalue(stats_within_burn_exp1_control$p[2])`) with no significant interaction between them $\beta$ = `r stats_within_burn_exp1_control$beta[5]`, `r make_pvalue(stats_within_burn_exp1_control$p[5])`.

The second experiment showed no difference that the cold intensity was less when the cold probe was located more distally for both control and TGI trials (TGI: $\beta$ = `r stats_within_cold_exp2_TGI$beta[2]`, `r make_pvalue(stats_within_cold_exp2_TGI$p[2])`; control: $\beta$ = `r stats_within_cold_exp2_control$beta[2]`, `r make_pvalue(stats_within_cold_exp2_control$p[2])`) with no significant interaction between them $\beta$ = `r stats_within_cold_exp2_control$beta[5]`, `r make_pvalue(stats_within_cold_exp2_control$p[5])`.
No difference in intensity was found for the warm ratings (TGI: $\beta$ = `r stats_within_warm_exp2_TGI$beta[2]`, `r make_pvalue(stats_within_warm_exp2_TGI$p[2])`; control: $\beta$ = `r stats_within_warm_exp2_control$beta[2]`, `r make_pvalue(stats_within_warm_exp2_control$p[2])`) with no significant interaction between them $\beta$ = `r stats_within_warm_exp2_control$beta[5]`, `r make_pvalue(stats_within_warm_exp2_control$p[5])`.
No significant differences was also found for the burning ratings for both TGI and control.
(TGI: $\beta$ = `r stats_within_burn_exp2_TGI$beta[2]`, `r make_pvalue(stats_within_burn_exp2_TGI$p[2])`; control: $\beta$ = `r stats_within_burn_exp2_control$beta[2]`, `r make_pvalue(stats_within_burn_exp2_control$p[2])`) with no significant interaction between them $\beta$ = `r stats_within_burn_exp2_control$beta[5]`, `r make_pvalue(stats_within_burn_exp2_control$p[5])`.

```{r, results="hide", fig.width = 15, fig.height = 15, warning = F, message = F, echo = F, fig.cap = "Difference in VAS ratings between the two dermatome conditions. The top panel shows the difference between proximal and distal cold afferent locations (within dermatomes) and the bottom panel shows the difference between rostral and caudal cold afferent locations, relative to warm, across stimulation type (Non-TGI, TGI) and VAS rating quality (cold, warm and burning). Small points show data from each participant, large dots are means across trials for each condition, and error bars show 95% confidence intervals."}
plot2()
```

@@

A main objective of this study was assessing spatial order effects along the rostrocaudal axis at the spinal level.
We delivered an equal number of trials in which the cold stimulus was applied on a dermatome that mapped more rostrally or caudally compared to the warm or neutral stimuli.

Firstly the cold intensity was more when the cold probe was located more rostrally for both control and TGI trials (TGI: $\beta$ = `r stats_across_cold_exp1_TGI$beta[2]`, `r make_pvalue(stats_across_cold_exp1_TGI$p[2])`; control: $\beta$ = `r stats_across_cold_exp1_control$beta[2]`, `r make_pvalue(stats_across_cold_exp1_control$p[2])`) with no significant interaction between them $\beta$ = `r stats_across_cold_exp1_control$beta[5]`, `r make_pvalue(stats_across_cold_exp1_control$p[5])`.
A significant reduction in warm intensity was found for TGI trials, but not for the control trials (TGI: $\beta$ = `r stats_across_warm_exp1_TGI$beta[2]`, `r make_pvalue(stats_across_warm_exp1_TGI$p[2])`; control: $\beta$ = `r stats_across_warm_exp1_control$beta[2]`, `r make_pvalue(stats_across_warm_exp1_control$p[2])`) with no significant interaction between them $\beta$ = `r stats_across_warm_exp1_control$beta[5]`, `r make_pvalue(stats_across_warm_exp1_control$p[5])`.
No significant differences was found for the burning ratings.(TGI: $\beta$ = `r stats_across_burn_exp1_TGI$beta[2]`, `r make_pvalue(stats_across_burn_exp1_TGI$p[2])`; control: $\beta$ = `r stats_across_burn_exp1_control$beta[2]`, `r make_pvalue(stats_across_burn_exp1_control$p[2])`) with no significant interaction between them ($\beta$ = `r stats_across_burn_exp1_control$beta[5]`, `r make_pvalue(stats_across_burn_exp1_control$p[5])`).

The second experiment showed that the cold intensity was more when the cold probe was located more rostrally for TGI trials, but not control trials.
(TGI: $\beta$ = `r stats_across_cold_exp2_TGI$beta[2]`, `r make_pvalue(stats_across_cold_exp2_TGI$p[2])`; control: $\beta$ = `r stats_across_cold_exp2_control$beta[2]`, `r make_pvalue(stats_across_cold_exp2_control$p[2])`) with a significant interaction between them $\beta$ = `r stats_across_cold_exp2_control$beta[5]`, `r make_pvalue(stats_across_cold_exp2_control$p[5])`.
A significant difference in intensity was found for the warm ratings indicating that when the cold probe was located more rostrally the intensity was lower (TGI: $\beta$ = `r stats_across_warm_exp2_TGI$beta[2]`, `r make_pvalue(stats_across_warm_exp2_TGI$p[2])`; control: $\beta$ = `r stats_across_warm_exp2_control$beta[2]`, `r make_pvalue(stats_across_warm_exp2_control$p[2])`) with no significant interaction between them $\beta$ = `r stats_across_warm_exp2_control$beta[5]`, `r make_pvalue(stats_across_warm_exp2_control$p[5])`.
A significant differences was also found for the burning ratings in the control condition, but not in the TGI condition, indicating more distally located cold probe decreased ratings in the control condition.
(TGI: $\beta$ = `r stats_across_burn_exp2_TGI$beta[2]`, `r make_pvalue(stats_across_burn_exp2_TGI$p[2])`; control: $\beta$ = `r stats_across_burn_exp2_control$beta[2]`, `r make_pvalue(stats_across_burn_exp2_control$p[2])`) with no significant interaction between them ($\beta$ = `r stats_across_burn_exp2_control$beta[5]`, `r make_pvalue(stats_across_burn_exp2_control$p[5])`).

>>>>

---
title: "Disentangling the spinal mechanisms of illusory heat and burning sensations in the Thermal Grill Illusion"
author: 
  - name : "**Jesper Fischer Ehmsen**"
    affiliation: "1"
  - name : "Daniel Elmstrom Christensen"
    affliation : "2"
affliation:

  - id : "1"
    institution : "Center of Functionally Integrative Neuroscience (CFIN), Department of Clinical Medicine, Aarhus University"
  - id : "2"
    institution : "Interacting Minds Center (IMC), Department XXX, Aarhus University"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
  html_document: default
  word_document:
bibliography: knitting files/Citations.bib
link-citations: yes
linkcolor: blue
always_allow_html: yes
editor_options: 
  markdown: 
    wrap: sentence
---

@@

The Thermal Grill Illusion (TGI) is a phenomenon in which the juxtaposition of innocuous warm and cold temperatures elicits a burning sensation, offering a unique window to understand how pain can be perceived in response to harmless stimuli.
Much debate has revolved around whether spinal mechanisms are involved in the generation of illusory pain, beyond supraspinal mechanisms.
In this study, we investigated the role of the spinal cord in the generation of the TGI, in two independent experiments, involving a total of 80 healthy individuals.
We applied heat and cold stimuli on dermatomes, namely areas of skin innervated by a single spinal nerve, mapped onto adjacent or nonadjacent spinal segments.
Participants were asked to rate their perceptions of cold, warm, and burning sensations in response to TGI and control stimuli.
Our aims were to investigate thermosensory and painful perceptual components of the TGI, as well as spatial features of the TGI that may illuminate processes underlying thermosensory integration in the spinal cord.
Our findings revealed that both thermosensory and painful components of TGI perception were modulated similarly, with enhanced warm and burning ratings observed when cold and warm stimuli were confined within the same dermatome.
Further, we found no perceptual differences based on the proximal-distal location of the cold stimulus within a single dermatome, but a notable enhancement of the TGI effect when the cold stimulus was associated with a more caudal segmental location along the spinal cord compared to the warm stimulus.
These results provide insights into the organisation of the spinal cord in relation to the thermosensory integration and generation of the TGI.

@@

The thermal grill illusion (TGI) is a perceptual phenomenon that challenges conventional understanding of pain perception.
It is a sensation of burning heat or pain when harmless cold and warm temperatures are applied to the skin simultaneously [@craig_thermal_1994; @craig_functional_1996].
Despite cold and warm temperatures being individually innocuous, their combination produces a contradictory burning sensation, even so the temperatures are insufficient to activate peripheral nociceptors.
The generation of this illusion is thus attributed to central nervous system mechanisms [@fardo_beyond_2020].
Recent studies have highlighted the involvement of the spinal cord as an initial site contributing to the TGI [@fardo_organization_2018; @harper_conditioned_2017].
However, the precise mechanisms underpinning integration of cold and warm thermal afferents in the spinal cord, alongside those responsible for the distinctive burning quality to this illusion, are yet to be elucidated.The TGI is often described as encompassing two distinct perceptual components - an illusion of heat and an illusion of pain [@defrin_spatial_2008, @fardo_beyond_2020]. The illusion of heat, also known as synthetic heat [@fruhstorfer_significance_2003; @green_localization_1977; -@green_synthetic_2002; -@green_temperature_2004], refers to non-painful sensations evoked by the thermal grill [@defrin_spatial_2008; @kern_pharmacological_2008; -@kern_effects_2008; @bouhassira_investigation_2005; @adam_relationships_2014].
The illusion of pain, which is the most recognised aspect of the TGI, is the distinctive burning sensation that accompanies the simultaneous presentation of cold and warm stimuli [@craig_thermal_1994; -@craig_functional_1996; @bach_thermal_2011].
The hallmark of both illusory components is a qualitative change in perception when the cold and warm stimuli are applied concurrently, compared to when they are presented individually.
Historically, the thermosensory and painful components of the TGI were explained by distinct spinal and supraspinal mechanisms, respectively.
The enhanced perception of heat in TGI was explained by a spinal inhibitory mechanism, drawing from observations in an animal model where simultaneous application of cold and warm temperatures reduced the activity of cold-specific spinal neurons compared to when cold was applied alone [@craig_thermal_1994]. Instead, the illusory pain component was ascribed to a disinhibition mechanism at the level of the thalamus, primarily based on the observations of unremitting pain following a thalamic lesion [@craig_thermal_1994; -@bud_craig_new_1998].
Recent human studies on TGI provided differing interpretations of the spinal or supraspinal origin of the illusion.
Two studies posited that the illusory pain of the TGI depends uniquely on supraspinal mechanisms.
This interpretation was based on the observed modulation of the illusion in accordance with a spatiotopic rather than somatotopic representation of the body [@marotta_transforming_2015].
Further, the illusion remained unaltered during concomitant tactile stimulation, suggesting ineffectiveness of tactile gating - a spinally-mediated process involving inhibition of nociceptive activity by concurrent somatosensory activity [@ferre_ineffectiveness_2018].
Counter to this perspective, other research endorsed a spinal contribution to the TGI.
These studies demonstrated that the illusion varied depending on whether cold and warm stimuli were applied to dermatomes mapped either onto adjacent or non-adjacent spinal segments [@fardo_organization_2018].
Participants perceived the stimulation more veridically, consistently with a reduction in TGI perception, when warm or cold stimuli triggered more widespread activity along the spinal cord, corroborating the hypothesis that the spinal cord is an initial site of thermosensory integration underlying TGI.
Further support for spinal mechanisms comes from research demonstrating that both noxious heat and the TGI were comparably reduced by conditioned pain modulation in humans.
This suggests a similar influence of descending modulation, irrespective of whether the painful sensation was triggered by potentially harmful (noxious) or harmless (innocuous) stimuli [@harper_conditioned_2017].
These findings collectively challenge a purely supraspinal hypothesis of the painful component of the TGI and indicate the relevance of spinal mechanisms in the manifestation of both illusory heat and pain within the TGI.
In this paper our objective was twofold.
Firstly, we directly investigated the hypothesis that thermosensory and burning components of the TGI experience are mediated by spinal mechanisms in humans, by manipulating the location of the stimuli within and across dermatomes.
Our past work using a similar manipulation involved measuring the experience of the TGI using a temperature matching task [@fardo_organization_2018], which provides a composite measure of TGI perception, reflecting both thermosensory and burning components.
Here, to probe possible distinctions between the two qualitative components of the TGI, we measured subjective indices of TGI perception using three independent visual analog scale (VAS) ratings of perceived cold, warm and burning sensations.
Secondly, we investigated spatial order effects associated with the integration of cold and warm sensory information at the dermatome (skin) and segmental (i.,e., spine) levels.
At the dermatome level, we used body-related coordinates to define proximal (towards the elbow) and distal (towards the wrist) locations.
At the spinal level, we used segment-related coordinates to define more rostral (towards the head) and more caudal (towards the lower back) locations.
Given the organisation of the spinal cord along a rostral-caudal axis, where each dermatome is represented across multiple spinal segments through the Lissauer tract, this study aimed to glean indirect insights into the spinal mechanisms underpinning thermosensory integration and the generation of the TGI.

@@

The Thermal Grill Illusion (TGI) is characterised by two key phenomena: thermosensory enhancement and illusory pain.
Thermosensory enhancement refers to an amplified perception of heat or cold when cold and warm stimuli are simultaneously applied, as opposed to when each stimulus is presented individually or paired with a neutral temperature.
Notably, the majority of individuals experience an intensification of heat rather than cold [@fardo_beyond_2020].
Illusory pain, on the other hand, denotes the perception of a burning sensation elicited by the pairing of warm and cold stimuli, an experience that is largely absent or significantly diminished when each stimulus is presented alone or combined with a neutral temperature.
Thus, indicators of a stronger TGI are heightened warm and burning ratings, coupled with reduced cold ratings.
To investigate thermosensory and burning components of the TGI, participants received pairs of temperatures on their forearms and were asked to quantify the levels of cold, warmth, and burning they experienced during each stimulation.
These stimuli consisted of either cold-warm pairs (TGI stimuli), which potentially evoked an illusion of heat and pain, or control stimuli that involved pairing a cold or a warm stimulus of identical temperature as in the TGI condition with a baseline temperature of 30°C (non-TGI stimuli).
All stimulation pairs were presented at a fixed distance on the skin, either within the same dermatome or across dermatomes that mapped onto non-adjacent spinal segments.
In doing so, the stimuli elicited cold-related and warm-related neural activity that was either more focal or more widespread along the spinal cord.
For each stimulation, participants reported their ratings using three sequential VAS scales ranging from 0, indicating the lack of a sensation, to 100, indicating an extreme sensation.
For each instance of stimulation, participants were guided to focus their reporting on the sensations originating from a specific thermode.
While the participants were unaware, this location corresponded to either the colder stimulus (Exp. 1) or the warmer stimulus (Exp. 2) of the paired temperatures.
VAS ratings were analysed using zero-inflated beta regressions.

@@

```{r, warning = F, message = F, echo = F,include=FALSE}
if (!require("pacman")) install.packages("pacman")
  pacman::p_load(boot, broom, broom.mixed, brms, car, caret, corrplot, DHARMa, gamlss,
                 ggeffects, gghalves, ggpol, ggpubr, glmmTMB, groupdata2, lme4, lmerTest, 
                 modelsummary, MuMIn, polycor, RColorBrewer, rcompanion, reshape2, Rmisc,
                 rsample, rstatix, simr, tibble, tidyverse, wesanderson, patchwork, cowplot
                 )
source(here::here("scripts","utility.R"))
source(here::here("scripts","plots.R"))

@@

In keeping with the unique heat and burning features of TGI, our results exhibited a more robust TGI when stimuli were confined within dermatomes compared to when applied across dermatomes, corresponding to non-adjacent spinal segments (Fig. 2).
We investigated this difference in subjective cold / warm and burning ratings with a two by two interaction of manipulation (control / TGI) and dermatome (within / across).
Results showed that participants reported a stronger reduced experience of cold, when the stimulus was presented within determatomes compared to across in the TGI condition compared to the control condition when the reference thermode was cold (experiment 1), but not when the reference thermode was warm (experiment 2)
(Exp. 1: $\beta$ = `r stats_model_cold_exp1$beta[6]`, `r make_pvalue(stats_model_cold_exp1$p[6])`; Exp. 2: $\beta$ = `r stats_model_cold_exp2$beta[6]`, `r make_pvalue(stats_model_cold_exp2$p[6])`)
The percept of warm showed no siginifcant interaction with either thermode as reference (Exp. 1: $\beta$ = `r stats_model_warm_exp1$beta[6]`, `r make_pvalue(stats_model_warm_exp1$p[6])`; Exp. 2: $\beta$ = `r stats_model_warm_exp2$beta[6]`, `r make_pvalue(stats_model_warm_exp2$p[6])`) 
The percept of burning showed no statistical significant interaction with the reference being cold, but a strong interaction when the reference thermode was warm
(Exp.1: $\beta$ = `r stats_model_burn_exp1$beta[6]`, `r make_pvalue(stats_model_burn_exp1$p[6])`; Exp. 2 $\beta$ = `r stats_model_burn_exp2$beta[6]`, `r make_pvalue(stats_model_burn_exp2$p[6])`)
These findings not only corroborated earlier results regarding the segmental organisation of the TGI [@fardo_organization_2018] but also suggested that both the thermosensory enhancement and illusory pain components of the illusion are modulated at the spinal cord level.
These findings are consistent with previous results on the modulation of heat and pain ratings of TGI stimuli by conditioned pain modulation [@harper_conditioned_2017].
All together these results support the role of spinal processes in the generation of distinct perceptual aspects of the TGI.

@@

Previous research demonstrated a phenomenon known as distal inhibition, wherein heat pain ratings tend to increase when a participant evaluates a more distal compared to a more proximal stimulus among two warm stimuli presented on the forearm [@quevedo_psychophysicshyperalgesia_2004; @quevedo_illusion_2007].
In our study, we sought to investigate whether this effect of distal inhibition would also influence TGI perception within a single dermatome (Fig. 3).
We tested this effect with a two way interaction between condition (control / TGI) and the location of the cold thermode when it was placed within dermatome (distal / proximal)
Results showed no statistical significant interaction effect for the cold ratings in both experiments:
(Exp. 1: $\beta$ = `r stats_model_cold_exp1$beta[8]`, `r make_pvalue(stats_model_cold_exp1$p[8])`; Exp. 2: $\beta$ = `r stats_model_cold_exp2$beta[8]`, `r make_pvalue(stats_model_cold_exp2$p[8])`)
Neither a statistically significant effect for warm watings
(Exp1: $\beta$ = `r stats_model_warm_exp1$beta[8]`, `r make_pvalue(stats_model_warm_exp1$p[8])`; Exp 2: $\beta$ = `r stats_model_warm_exp2$beta[8]`, `r make_pvalue(stats_model_warm_exp2$p[8])`) 
and also not for burning ratings
(Exp1: $\beta$ = `r stats_model_burn_exp1$beta[8]`, `r make_pvalue(stats_model_burn_exp1$p[8])`; Exp 2: $\beta$ = `r stats_model_burn_exp2$beta[8]`, `r make_pvalue(stats_model_burn_exp2$p[8])`)

```{r, results="hide", fig.width = 7.2, fig.height = 7.2, warning = F, message = F, echo = F, fig.cap = "Difference in VAS ratings between the two dermatome conditions. The top panel shows the difference between proximal and distal cold afferent locations (within dermatomes) and the bottom panel shows the difference between rostral and caudal cold afferent locations, relative to warm, across stimulation type (Non-TGI, TGI) and VAS rating quality (cold, warm and burning). Small points show data from each participant, large dots are means across trials for each condition, and error bars show 95% confidence intervals."}
plot2()
```

@@

A main objective of this study was assessing spatial order effects along the rostrocaudal axis at the spinal level.
We delivered an equal number of trials in which the cold stimulus was applied on a dermatome that mapped more rostrally or caudally compared to the warm or neutral stimuli.
Like the previous analysis we investigated this relationship with a two way interaction here with the rostral - caudal axis being investigated.
Results showed no statistical significant interaction effect for the cold ratings in the first experiment, but a statistically significant interaction when the reference thermode was warm:
(Exp. 1: $\beta$ = `r stats_model_cold_exp1$beta[7]`, `r make_pvalue(stats_model_cold_exp1$p[7])`; Exp. 2: $\beta$ = `r stats_model_cold_exp2$beta[7]`, `r make_pvalue(stats_model_cold_exp2$p[7])`)
No statistically significant effect for warm watings
(Exp1: $\beta$ = `r stats_model_warm_exp1$beta[7]`, `r make_pvalue(stats_model_warm_exp1$p[7])`; Exp 2: $\beta$ = `r stats_model_warm_exp2$beta[7]`, `r make_pvalue(stats_model_warm_exp2$p[7])`) 
and also not for burning ratings
(Exp1: $\beta$ = `r stats_model_burn_exp1$beta[7]`, `r make_pvalue(stats_model_burn_exp1$p[7])`; Exp 2: $\beta$ = `r stats_model_burn_exp2$beta[7]`, `r make_pvalue(stats_model_burn_exp2$p[7])`)
